---
title: "Hands-on Exercise 1"
author: "Ng Meng Ye"
---

# Getting Started

## Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages have been installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## R Graphics VS ggplot

```{r}
#| code-fold: true
#| code-summary: "Show the code"
hist(exam_data$MATHS)
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
ggplot(data=exam_data,
       aes(x = MATHS)) + 
  geom_histogram(bins=10,
                 boundary = 100,
                 color = "black",
                 fill="grey")+
  ggtitle("Distribution of Maths scores")
```

::: callout-note
Why ggplot2 is recommended? ggplot2 looks better
:::

# Essential Grammatical Elements in ggplot2:data

```{r}
ggplot(data=exam_data)
```

::: callout-note
-   A blank canvas appears.
-   ggplot() initializes a ggplot object.
-   The data argument defines the dataset to be used for plotting.
-   If the dataset is not already a data.frame, it will be converted to one by fortify().
:::

```{r}
ggplot(data=exam_data,
       aes(x=MATHS))
```
#Essential Grammatical Elements in ggplot2: geom

## geom_bar

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```
```{r}
ggplot(data=exam_data, 
       aes(x=MATHS)) +
  geom_boxplot()
```
## geom_dotplot

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```
## geom_histogram()

