---
title: "Hands-on Exercise 1"
author: "Ng Meng Ye"
---

# Getting Started

## Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages have been installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## R Graphics VS ggplot

```{r}
#| code-fold: true
#| code-summary: "Show the code"
hist(exam_data$MATHS)
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
ggplot(data=exam_data,
       aes(x = MATHS)) + 
  geom_histogram(bins=10,
                 boundary = 100,
                 color = "black",
                 fill="grey")+
  ggtitle("Distribution of Maths scores")
```

::: callout-note
Why ggplot2 is recommended? ggplot2 looks better
:::

# Essential Grammatical Elements in ggplot2:data

```{r}
ggplot(data=exam_data)
```

::: callout-note
-   A blank canvas appears.
-   ggplot() initializes a ggplot object.
-   The data argument defines the dataset to be used for plotting.
-   If the dataset is not already a data.frame, it will be converted to one by fortify().
:::

```{r}
ggplot(data=exam_data,
       aes(x=MATHS))
```
# Essential Grammatical Elements in ggplot2: geom

## geom_bar

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```
```{r}
ggplot(data=exam_data, 
       aes(x=MATHS)) +
  geom_boxplot()
```
## geom_dotplot

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```
## geom_histogram


```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()       

```


## Modify geom()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")  

```

## Modify aes()


```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")

```

## geom_density

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()           
```

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()

```

## geom_boxplot

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()            

```

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```
## geom_violin


```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

## geom_point

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```

## Combine geom objects

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```
# Essential Grammatical Elements in ggplot2: stat

## stat

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

## stat_summary

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)               
```

## Add best fit curve on a scatterplot

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```
# Essential Grammatical Elements in ggplot2: Facets


## facet_wrap


```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)

```

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

# Essential Grammatical Elements in ggplot2: Coordinates

## Coordinate

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```
## Change y and x axis range

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```
# Essential Grammatical Elements in ggplot2: themes

## theme

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```


