---
title: "Hands-on Exercise 2"
author: "Ng Meng Ye"
---

# Getting Started

## Install and launching the required libraries

### Local Package Check

```{r}
#| echo: false
#| code-fold: true
#| code-summary: "Show the code"
pkg_list <- data.frame(
  Package = c("tidyverse", "ggrepel", "patchwork", "ggthemes", "hrbrthemes"),
  Status = c("Data manipulation and core visualization", "Prevents text labels from overlapping on plots", "Combines multiple plots into one layout", "Premade themes", "igh-quality, typography-centric themes")
)

knitr::kable(pkg_list, caption = "Packages for this Project")
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse) 
```

## Importing Data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

```{r}
glimpse(exam_data)
```

```{r}
summary(exam_data)
```

## Beyond ggplot2 Annotation: ggrepel

### Without ggrepel

::: panel-tabset
### The plot

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

### The code

``` r
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

### Working with ggrepel

::: panel-tabset
### The plot

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

### The code

``` r
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

## Beyond ggplot2 Themes

::: panel-tabset
### The plot

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_light() +
  ggtitle("Distribution of Maths scores") 

```

### The code

``` r
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```
:::

### Working with ggtheme package

::: panel-tabset
### The plot

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_economist() + 
  theme(
    # Sets the background to the "Economist" cream color
    panel.background = element_rect(fill = "#E9E4E0", color = NA),
    plot.background = element_rect(fill = "#E9E4E0", color = NA),
    
    # Changes the horizontal grid lines from white to light grey
    panel.grid.major.y = element_line(color = "grey80"), 
    
    # Ensures vertical grid lines are removed (typical for this style)
    panel.grid.major.x = element_blank()
  ) +
  ggtitle("Distribution of Maths scores")

```

### The code

``` r
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_economist() +
  ggtitle("Distribution of Maths scores") 
```
:::

### Working with hrbthemes package

::: panel-tabset
### The plot

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

### The code

``` r
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
### The code
```
:::

::: panel-tabset
### The plot

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

### The code

``` r
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```
:::

::: callout-note
-   `axis_title_size` argument is used to increase the font size of the axis title to 18

-   `base_size` argument is used to increase the default axis label to 15

-   `grid` argument is used to remove the x-axis grid lines
:::

## Beyond Single Graph

::: panel-tabset
### The plot

```{r}
#| echo: false
#| message: false
#| warning: false
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")

p1
```

### The code

``` r
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")
```
:::

::: panel-tabset
### The plot

```{r}
#| echo: false
#| message: false
#| warning: false
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")

p2
```

### The code

``` r
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")
```
:::

::: panel-tabset
### The plot

```{r}
#| echo: false
#| message: false
#| warning: false
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")

p3
```

### The code

``` r
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

### Creating Composite Graphics

### Combining 2 ggplot2 graphs

::: panel-tabset
### The plot

```{r}
#| echo: false
#| message: false
#| warning: false
p1 + p2
```

### The code

``` r
p1 + p2
```
:::

### Combining 3 ggplot2 graphs

::: panel-tabset
### The plot

```{r}
#| echo: false
#| message: false
#| warning: false
(p1 / p2) | p3
```

### The code

``` r
(p1 / p2) | p3
```
:::

### Creating a composite figure with tag

::: panel-tabset
### The plot

```{r}
#| echo: false
#| message: false
#| warning: false
((p1 / p2) | p3)+ 
  plot_annotation(tag_levels = 'I')
```

### The code

``` r
((p1 / p2) | p3)+ 
  plot_annotation(tag_levels = 'I')
```
:::

### Creating figure with insert

::: panel-tabset
### The plot

```{r}
#| echo: false
#| message: false
#| warning: false
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

### The code

``` r
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```
:::

### Creating a composite figure by using patchwork and ggtheme


```{r}
p1 + p2
```


::: panel-tabset
### The plot

```{r}
patchwork <- (p1 / p2) | p3
patchwork
```

### The code

``` r
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```
:::



