---
title: "COFINFAD: Colombian Fintech Financial Analysis"
author: "Ng Meng Ye"
---

# Project Overview & Background

This project examines the **digital service adoption patterns** of 48,723 customers within a Colombian fintech ecosystem, analyzing **3,159,157 transactions** from January to December 2023.

## Regional Context: Colombia's Digital Pivot

Colombia serves as a "global testing ground" for financial innovation due to its unique socioeconomic landscape:

-   **Demographics**: With a population of approximately **54 million** (2026 est.), Colombia has a relatively young and urbanized demographic. The median age is **33 years**, and over **80%** of the population resides in urban centers like Bogotá, Medellín, and Cali. This young, tech-savvy population is a primary driver for mobile-first financial services.

-   **Economic Profile (GDP)**: As an upper-middle-income economy, Colombia's GDP is projected at approximately **\$419 billion (2024)**, with a GDP per capita of around **\$7,920**. While broader economic growth hovers around 2–3%, the **digital commerce sector** is outpacing the general economy, expanding at double-digit rates (approx. 12–15% annually).

-   **Financial Inclusion**: Historically, large segments of the Colombian population were "underbanked" due to geographic and cost barriers. However, digital banking has surged; as of 2024, nearly **78% of transactions** in the country are digital. This shift is supported by the National Digital Strategy (2023–2026), which promotes interoperable instant-payment frameworks.

## Our Data

The dataset captures this transition from traditional cash-based behavior to a digital-first economy.

-   **Transactional Data**: Captures the "pulse" of activity across 3.1M+ records, categorized into `Transfer`, `Withdrawal`, and `Payment`.

-   **Customer Profiles**: Includes 54 variables that allow us to see how demographics (like the **income_bracket** or **location**) correlate with **app_logins_frequency** and **satisfaction_score**.

## Data Preparation

Given the scale of the Colombian dataset (116MB CSV), we prepare the data for efficient R-based analysis:

1.  **Compression**: Raw `.csv` files are converted to `.rds` format to fit within GitHub's 100MB limit while maintaining high-speed loading in R.

2.  **Relational Integration**: We join the transactional "Fact" table with the customer "Dimension" table using `customer_id`.

3.  **Variable Encoding**: Categorical metrics (e.g., `acquisition_channel`, `clv_segment`) are converted to factors to facilitate the **Confirmatory Data Analysis (CDA)** phase.

```{r}
library(tidyverse)

# 1. Load the raw data
transactions_raw <- read_csv("data/transactions_data.csv")
customer_raw <- read_csv("data/customer_data.csv")

# 2. Save as RDS (This will likely shrink the 116MB file to <25MB)
saveRDS(transactions_raw, "data/transactions_data.rds")
saveRDS(customer_raw, "data/customer_data.rds")

# Load the lightweight RDS files
transactions <- readRDS("data/transactions_data.rds")
customers <- readRDS("data/customer_data.rds")

# Prepare Master Data Frame
master_df <- transactions %>%
  mutate(date = as.Date(date, format = "%m/%d/%Y")) %>%
  left_join(customers, by = "customer_id")
```

# The "Pulse" of the Fintech (Temporal Analysis)

Before looking at individuals, look at the ecosystem. Since your data covers all of 2023, you want to see the "heartbeat" of Colombian digital finance.

-   **What to do:** Plot daily and monthly transaction volumes.

-   **Observation:** Look for "Payday Peaks" (usually the 15th and 30th of the month in Colombia) and holiday surges (December).

-   **Tech Suggestion:** Use `ggplotly()` on a line chart so you can hover over specific days to see the exact transaction count.

## Daily Pulse (Line Chart)

```{r}
# 1. Load Libraries
# Load required libraries
library(tidyverse)
library(lubridate)

# 1. Load Data
transactions <- readRDS("data/transactions_data.rds")

# 2. Prep Daily Data
daily_vol <- transactions %>%
  mutate(date = as.Date(date, format = "%m/%d/%Y")) %>%
  group_by(date) %>%
  summarise(tx_count = n())

# 3. Plot Daily Pulse
ggplot(daily_vol, aes(x = date, y = tx_count)) +
  geom_line(color = "#2c3e50", size = 0.5) +
  # Adding a trend line to see the big picture
  geom_smooth(method = "loess", span = 0.1, color = "#e74c3c", se = FALSE) +
  labs(title = "Daily Transaction Pulse (2023)",
       subtitle = "The red line indicates the 12-month moving trend",
       x = "Month", y = "Transaction Count") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  theme_minimal()
```

## Average Volume by Day of Month (Bar Chart)

```{r}
# 1. Prep Day-of-Month Data
# First, get the count for every unique day in the calendar
library(tidyverse)
library(lubridate)

# 1. Load Data
transactions <- readRDS("data/transactions_data.rds")

# 2. Preparation (Normalization Logic)
day_counts <- transactions %>%
  mutate(date = as.Date(date, format = "%m/%d/%Y"),
         day_of_month = mday(date)) %>%
  group_by(date, day_of_month) %>%
  summarise(daily_total = n(), .groups = "drop")

avg_payday_pulse <- day_counts %>%
  group_by(day_of_month) %>%
  summarise(avg_transactions = mean(daily_total))

# 3. Plot with Zoomed Y-Axis
ggplot(avg_payday_pulse, aes(x = day_of_month, y = avg_transactions)) +
  geom_col(fill = "#3498db", color = "white") + 
  scale_x_continuous(breaks = 1:31) +
  # This line starts the view at 7,500 and lets the top adjust automatically
  coord_cartesian(ylim = c(5000, 8900)) +
  labs(title = "Average Daily Transaction Volume by Day of Month",
       x = "Day of Month", 
       y = "Mean Transactions per Day") +
  theme_minimal()
```

# Demographic & Geographic Adoption

With 48,000+ customers across cities like Bogotá, Medellín, and Cali, we need to see who is actually using the app.

-   **Treemap:** Transaction volume by `location`. (Is Bogotá dominating, or are smaller cities like Pereira showing higher growth per capita?)

## Top 10 Cities by User Base

```{r}
library(tidyverse)

# 1. Load Customer Data
customers <- readRDS("data/customer_data.rds")

# 2. Process Location Data
# Splitting "City, Department" and counting users
city_dist <- customers %>%
  separate(location, into = c("city", "department"), sep = ", ") %>%
  group_by(city) %>%
  summarise(user_count = n()) %>%
  arrange(desc(user_count)) %>%
  slice_head(n = 10)

# 3. Static Bar Chart
ggplot(city_dist, aes(x = reorder(city, user_count), y = user_count)) +
  geom_col(fill = "#2ecc71", color = "white") +
  coord_flip() + # Flip for better readability of city names
  labs(title = "Top 10 Colombian Cities by User Base",
       subtitle = "Concentration of the 48,723 unique customers",
       x = "City", 
       y = "Number of Customers") +
  theme_minimal()
```

## Treemap of Customer Distribution

```{r}
pacman::p_load(treemap,treemapify)
```

```{r}

library(tidyverse)
library(treemap)
library(treemapify)

geo_data <- customers %>%
  separate(location, into = c("city", "department"), sep = ", ") %>%
  group_by(department, city) %>%
  summarise(user_count = n(), .groups = "drop")

# 3. Create the Treemap
# Following the 'treemap' package syntax from R4VA
treemap(geo_data,
        index = c("department", "city"), # Hierarchy: Dept first, then City
        vSize = "user_count",            # Size based on number of users
        type = "index",                  # Color based on index levels
        title = "Geographic Distribution of Customers in Colombia",
        palette = "Set3",                # Professional color palette
        border.col = c("black", "white"),# Distinct borders for levels
        border.lwds = c(3, 1),           # Thicker borders for Departments
        fontsize.labels = c(12, 8),      # Larger font for Dept, smaller for City
        fontcolor.labels = "black",
        inflate.labels = FALSE)
```
