---
title: "COFINFAD: Colombian Fintech Financial Analysis"
author: "Ng Meng Ye"
---

# Project Overview & Background

## The Objective

The objective of this project is to understand the **behaviors behind customer transactions** and how **socio-demographic factors**—such as age, income, and occupation—affect customer transactions. By applying statistical correlation techniques, we aim to identify which personal attributes are the true drivers of financial activity within the Colombian fintech ecosystem.

## Regional Context: Socio-Demographic Diversity

Colombia’s unique social structure provides a rich environment for demographic behavioral analysis:

-   **Urban-Rural Divide**: Economic activity is heavily concentrated in the "Golden Triangle" of **Bogotá, Medellín, and Cali**. We investigate if geography is a primary factor in transaction frequency or if digital adoption has bridged the regional gap.

-   **The Emerging Middle Class**: With a median age of **33**, the Colombian workforce is transitioning into digital-first banking. This project evaluates how **income brackets** and **education levels** correlate with transaction volumes to see if wealth is a prerequisite for high engagement.

-   **Professional Variance**: From independent freelancers to corporate engineers, we analyze how **occupational status** acts as a factor in how frequently users interact with their digital wallets.

## Analytical Framework: Socio-Demographic Correlation

To isolate the influence of these factors, we utilize a **Correlation-First** approach:

1.  **Metric Derivation**: We normalize the data by calculating the **Average Transaction Value** (Total Volume / Count) to move beyond surface-level totals and understand the "per-action" behavior of different groups.

2.  **Non-Parametric Testing**: Recognizing that income and transaction data are often heavily skewed, we employ **Spearman’s Rank Correlation** (via `ggstatsplot`) to assess the strength of relationships between demographic variables and transaction intensity.

3.  **Faceted Comparison**: We utilize faceting to determine if socio-demographic correlations remain consistent across different segments, or if certain groups (e.g., specific age brackets or genders) exhibit unique behavioral outliers.

## Our Data

The dataset captures this transition from traditional cash-based behavior to a digital-first economy.

-   **Transactional Data**: Captures the "pulse" of activity across 3.1M+ records, categorized into `Transfer`, `Withdrawal`, and `Payment`.

-   **Customer Profiles**: Includes 54 variables that allow us to see how demographics (like the **income_bracket** or **location**) correlate with **app_logins_frequency** and **satisfaction_score**.

## Data Preparation

Given the scale of the Colombian dataset (116MB CSV), we prepare the data for efficient R-based analysis:

1.  **Compression**: Raw `.csv` files are converted to `.rds` format to fit within GitHub's 100MB limit while maintaining high-speed loading in R.

2.  **Relational Integration**: We join the transactional "Fact" table with the customer "Dimension" table using `customer_id`.

3.  **Variable Encoding**: Categorical metrics (e.g., `acquisition_channel`, `clv_segment`) are converted to factors to facilitate the **Confirmatory Data Analysis (CDA)** phase.

```{r}
#| warning: false
#| message: false
library(tidyverse)

# 1. Load the raw data
transactions_raw <- read_csv("data/transactions_data.csv")
customer_raw <- read_csv("data/customer_data.csv")

# 2. Save as RDS (This will likely shrink the 116MB file to <25MB)
saveRDS(transactions_raw, "data/transactions_data.rds")
saveRDS(customer_raw, "data/customer_data.rds")

# Load the lightweight RDS files
transactions <- readRDS("data/transactions_data.rds")
customers <- readRDS("data/customer_data.rds")

# Prepare Master Data Frame
master_df <- transactions %>%
  mutate(date = as.Date(date, format = "%m/%d/%Y")) %>%
  left_join(customers, by = "customer_id")
```

The code chunk below will be used to install and load these packages in RStudio.

```{r}
#| warning: false
#| message: false
pacman::p_load("ggstatsplot", "plotly", "DT", "scales","tidyverse")
```

# Correlation

## Analyzing how different categories affects the Transaction Amount vs. Frequency behavior

### 1. Education Level

The function 'a' is the socio-demographic factor (e.g., "gender", "income_bracket")

```{r}
options(scipen = 999)

plot_behavior_correlation <- function(a) {
  ggscatterstats(
    data = customers, 
    x = avg_tx_value,  
    y = tx_count,
    type = "nonparametric", 
    title = paste("Socio-Demographic Factor Analysis:", a),
    xlab = "Average Transaction $",
    ylab = "Transaction Frequency"
  ) + 
    facet_wrap(vars(!!sym(a))) +
    # This part cleans the x-axis
    scale_x_continuous(
      labels = label_number(scale = 1e-6, suffix = "M"),
      breaks = seq(0, 50000000, by = 10000000) # Optional: set breaks every 10M
    )
}
```

```{r}
#| warning: false
#| message: false
plot_behavior_correlation("education_level")
```

### 2. Marital Status

```{r}
#| warning: false
#| message: false
plot_behavior_correlation("marital_status")
```

### 3. Gender

```{r}
#| warning: false
#| message: false
plot_behavior_correlation("gender")
```

###  4. Age

```{r}
library(dplyr)

customers <- customers %>%
  mutate(age_group = case_when(
    age <= 35 ~ "20-35", # Assuming the youngest users start around 20
    age >= 36 & age <= 50 ~ "36-50",
    age >= 51 & age <= 65 ~ "51-65",
    age > 65  ~ "65+",
    TRUE      ~ "Others"
  )) %>%
  # Re-ordering the factors for the plot
  mutate(age_group = factor(age_group, 
                            levels = c("20-35", "36-50", "51-65", "65+")))
```

```{r}
#| warning: false
#| message: false
plot_behavior_correlation("age_group")
```

### 5. Income Bracket

```{r}
#| warning: false
#| message: false
plot_behavior_correlation("income_bracket")

```

### 6. Household Size

```{r}

customers <- customers %>%
  mutate(hh_group = case_when(
    household_size == 1 ~ "1",
    household_size >= 2 & household_size <= 3 ~ "2-3",
    household_size >= 4 & household_size <= 5 ~ "4-5",
    household_size >= 6 ~ "6+",
    TRUE ~ "Unknown"
  )) %>%
  # Set levels to ensure the plot follows the correct order
  mutate(hh_group = factor(hh_group, 
                           levels = c("1", "2-3", "4-5", "6+")))

```

```{r}
#| warning: false
#| message: false
plot_behavior_correlation("hh_group")

```
