---
title: "COFINFAD: Colombian Fintech Financial Analysis"
author: "Ng Meng Ye"
---

# Project Overview & Background

## The Objective

The objective of this project is to understand the **behaviors behind customer transactions** and how **socio-demographic factors**—such as age, income, and occupation—affect customer transactions. By applying statistical correlation techniques, we aim to identify which personal attributes are the true drivers of financial activity within the Colombian fintech ecosystem.

## Analytical Framework: Socio-Demographic Correlation

To isolate the influence of these factors, we utilize a **Correlation-First** approach:

1.  **Metric Derivation**: We normalize the data by calculating the **Average Transaction Value** (Total Volume / Count) to move beyond surface-level totals and understand the "per-action" behavior of different groups.

2.  **Non-Parametric Testing**: Recognizing that income and transaction data are often heavily skewed, we employ **Spearman’s Rank Correlation** (via `ggstatsplot`) to assess the strength of relationships between demographic variables and transaction intensity.

3.  **Faceted Comparison**: We utilize faceting to determine if socio-demographic correlations remain consistent across different segments, or if certain groups (e.g., specific age brackets or genders) exhibit unique behavioral outliers.

## Our Data

The dataset captures this transition from traditional cash-based behavior to a digital-first economy.

-   **Customer Profiles**: Includes 54 variables that allow us to see how demographics (like the **income_bracket** or **location**) correlate with **tx_count** and **avg_tx_value**.

## Data Preparation

Raw `.csv` files are converted to `.rds` format to fit within GitHub's 100MB limit while maintaining high-speed loading in R.

```{r}
#| warning: false
#| message: false
library(tidyverse)

customer_raw <- read_csv("data/customer_data.csv")

saveRDS(customer_raw, "data/customer_data.rds")

customers <- readRDS("data/customer_data.rds")

```

The code chunk below will be used to install and load these packages in RStudio.

```{r}
#| warning: false
#| message: false
pacman::p_load(ggstatsplot, plotly, DT, scales,tidyverse,gtsummary,ggdist, ggridges, ggthemes,colorspace)
```

# Exploratory Data Analysis

## Data Exploration

```{r}
library(tidyverse)
glimpse(customers)
```

## Data Cleaning

We only keep those columns that are required for our analysis

1.  customer_id
2.  age
3.  gender
4.  location
5.  income_bracket
6.  education_level
7.  household_size
8.  marital_status
9.  tx_count
10. avg_tx_volume
11. total_tx_volume

```{r}
library(tidyverse)

# Retaining only the specified columns
customers_cleaned <- customers %>%
  select(
    customer_id,
    age,
    gender,
    location,
    income_bracket,
    education_level,
    household_size,
    marital_status,
    tx_count,
    avg_tx_value,
    total_tx_volume
  )

# Verify the result
glimpse(customers_cleaned)
```

Create age group

```{r}

customers_cleaned <- customers_cleaned %>%
  mutate(age_group = case_when(
    age <= 35 ~ "20-35", # Assuming the youngest users start around 20
    age >= 36 & age <= 50 ~ "36-50",
    age >= 51 & age <= 65 ~ "51-65"
  )) %>%
  # Re-ordering the factors for the plot
  mutate(age_group = factor(age_group, 
                            levels = c("20-35", "36-50", "51-65")))
```

Create household_size group

```{r}

customers_cleaned <- customers_cleaned %>%
  mutate(hh_group = case_when(
    household_size == 1 ~ "1",
    household_size >= 2 & household_size <= 3 ~ "2-3",
    household_size >= 4 & household_size <= 5 ~ "4-5",
    household_size >= 6 ~ "6+",
    TRUE ~ "Unknown"
  )) %>%
  # Set levels to ensure the plot follows the correct order
  mutate(hh_group = factor(hh_group, 
                           levels = c("1", "2-3", "4-5", "6+")))
```

## Descriptive Statistic

Review the statistical summary of the **socio-demographic factors**

```{r}
library(gt)

descriptive_summary_table <- customers_cleaned %>%
  select(education_level, marital_status, gender, age_group, 
         income_bracket, hh_group) %>%
  tbl_summary(
    # ... (keep your existing tbl_summary code here)
  ) %>%
  as_gt() %>% # Convert gtsummary to gt object
  tab_header(
    title = "Socio-Demographic Descriptive Statistics",
    subtitle = "Colombian Fintech User Base (2023)"
  ) %>%
  opt_stylize(color = "blue", style = 1) # Built-in themes

descriptive_summary_table
```

Review the statistical summary of the **spending behaviours**

```{r}
library(gtsummary)
library(gt)

continuous_summary_table <- customers_cleaned %>%
  select(tx_count, avg_tx_value, total_tx_volume) %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd}) | {median} ({p25}, {p75}) | [{min}, {max}]"
    ),
    digits = all_continuous() ~ 0, # Financial volumes don't need decimals
    label = list(
      tx_count ~ "Transaction Count (Frequency)",
      avg_tx_value ~ "Average Transaction Value (COP)",
      total_tx_volume ~ "Total Transaction Volume (COP)"
    )
  ) %>%
  modify_header(label = "**Financial Metric**") %>%
  as_gt() %>%
  tab_header(
    title = "Financial Performance Descriptive Statistics",
    subtitle = "Summary of User Transaction Behavior (2023)"
  ) %>%
  opt_stylize(color = "blue", style = 1)

continuous_summary_table
```

```{r}
ggplot(data = customers_cleaned, 
       aes(x = avg_tx_value)) +
  geom_histogram(bins = 20, 
                 boundary = 100,
                 color = "grey25", 
                 fill = "grey90") +
  # Use labs to define the x-axis title
  labs(x = "Average Transaction $", 
       title = "Distribution of Average Transaction $") +
  theme_economist() +
  # This line is CRITICAL: Economist theme hides titles by default
  theme(axis.title.x = element_text(display = "block", size = 12, vjust = -1))
```

Analyzing how different categories affects the Transaction Amount vs. Frequency behavior

1.  Education Level
2.  Marital Status
3.  Gender
4.  Age
5.  Income Bracket
6.  Household Size

### 1. Education Level

Visualizing distribution using half eye graph and boxplot

```{r}
plot_distribution <- function(cat_var) {
  
  ggplot(customers_cleaned,
         aes(x = {{cat_var}}, 
             y = avg_tx_value)) +
    
    stat_halfeye(
      adjust = 0.5,
      justification = -0.2,
      .width = 0,
      point_colour = NA
    ) +
    
    geom_boxplot(
      width = 0.20,
      outlier.shape = NA
    ) +
    
    stat_dots(
      side = "left",
      justification = 1.2,
      binwidth = 0.5,
      dotsize = 2
    ) +
    
    labs(
      x = deparse(substitute(cat_var)),
      y = "Average Transaction Value",
      title = paste("Distribution of Average Transaction Value by", 
                    deparse(substitute(cat_var)))
    ) +
    
    theme_minimal()
}
```

```{r}
plot_distribution(education_level)
```

Visualizing distribution using half eye graph and boxplot on a Log Scale (Base 10)

```{r}
plot_demo_log <- function(cat_var) {
  
  ggplot(customers_cleaned, 
         aes(x = {{cat_var}}, 
             y = avg_tx_value, 
             fill = {{cat_var}})) +
    
    stat_halfeye(
      adjust = 0.5,
      justification = -0.2,
      .width = 0,
      point_colour = NA
    ) +
    
    geom_boxplot(
      width = 0.20,
      outlier.shape = NA
    ) +
    
    scale_y_log10(
      labels = scales::label_number(scale = 1e-6, suffix = "M"),
      name = "Avg Transaction Value (COP) - Log Scale"
    ) +
    
    labs(
      title = paste("Socio-Demographic Analysis (Log-Transformed):",
                    deparse(substitute(cat_var))),
      x = deparse(substitute(cat_var))
    ) +
    
    theme_minimal()
}
```

```{r}
plot_demo_log(education_level)
```

Correlation Analysis (relationship between average transaction value vs transaction count (frequency)

```{r}
options(scipen = 999)

plot_behavior_correlation <- function(a) {
  ggscatterstats(
    data = customers_cleaned, 
    x = avg_tx_value,  
    y = tx_count,
    type = "nonparametric", 
    title = paste("Socio-Demographic Factor Analysis:", a),
    xlab = "Average Transaction Value",
    ylab = "Transaction Frequency"
  ) + 
    facet_wrap(vars(!!sym(a))) +
    # This part cleans the x-axis
    scale_x_continuous(
      labels = label_number(scale = 1e-6, suffix = "M"),
      breaks = seq(0, 50000000, by = 10000000) # Optional: set breaks every 10M
    )
}
```

```{r}
#| warning: false
#| message: false
plot_behavior_correlation("education_level")
```

### 2. Marital Status

Visualizing distribution using half eye graph and boxplot

```{r}
plot_distribution(marital_status)
```

Visualizing distribution using half eye graph and boxplot on a Log Scale (Base 10)

```{r}
plot_demo_log(marital_status)
```

```{r}
#| warning: false
#| message: false
plot_behavior_correlation("marital_status")
```

### 3. Gender

```{r}
plot_distribution(gender)
```

```{r}
plot_demo_log(gender)
```

```{r}
#| warning: false
#| message: false
plot_behavior_correlation("gender")
```

### 4. Age

```{r}
plot_distribution(age_group)
```

```{r}
plot_demo_log(age_group)
```

```{r}
#| warning: false
#| message: false
plot_behavior_correlation("age_group")


```

### 5. Income Bracket

```{r}
plot_distribution(income_bracket)
```

```{r}
plot_demo_log(income_bracket)
```

```{r}
#| warning: false
#| message: false
plot_behavior_correlation("income_bracket")
```

### 6. Household Size

```{r}
plot_distribution(hh_group)
```

```{r}
plot_demo_log(hh_group)
```

```{r}
#| warning: false
#| message: false
plot_behavior_correlation("hh_group")
```

# Confirmatory Data Analysis

## Oneway ANOVA Test: *ggbetweenstats()* method

In the code chunk below, [*ggbetweenstats()*](https://indrajeetpatil.github.io/ggstatsplot/reference/ggbetweenstats.html) is used to build a visual for One-way ANOVA test on Average Transaction Amount by different variables.

### 1. Education Level

```{r}

ggbetweenstats(
  data = customers_cleaned,
  x = education_level,
  y = avg_tx_value,
  type = "nonparametric",
  ylab = "Average Transaction Value",
  xlab = "Education Level",
  title = "Average Transaction Value by Education Level",
  mean.ci = TRUE, 
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

### 2. Marital Status

```{r}

ggbetweenstats(
  data = customers_cleaned,
  x = marital_status,
  y = avg_tx_value,
  type = "nonparametric",
  ylab = "Average Transaction Value",
  xlab = "Marital Status",
  title = "Average Transaction Value by Marital Status",
  mean.ci = TRUE, 
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)

```

### 3. Gender

```{r}

ggbetweenstats(
  data = customers_cleaned,
  x = gender,
  y = avg_tx_value,
  type = "nonparametric",
  ylab = "Average Transaction Value",
  xlab = "Gender",
  title = "Average Transaction Value by Gender",
  mean.ci = TRUE, 
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)


```

### 4. Age

```{r}

ggbetweenstats(
  data = customers_cleaned,
  x = age_group,
  y = avg_tx_value,
  type = "nonparametric",
  ylab = "Average Transaction Value",
  xlab = "Age Group",
  title = "Average Transaction Value by Age Group",
  mean.ci = TRUE, 
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

### 5. Income Bracket

```{r}
ggbetweenstats(
  data = customers_cleaned,
  x = income_bracket,
  y = avg_tx_value,
  type = "nonparametric",
  ylab = "Average Transaction Value",
  xlab = "Income Bracket",
  title = "Average Transaction Value by Income Bracket",
  mean.ci = TRUE, 
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

### 6. Household Size

```{r}
ggbetweenstats(
  data = customers_cleaned,
  x = hh_group,
  y = avg_tx_value,
  type = "nonparametric",
  ylab = "Average Transaction Value",
  xlab = "Household Group Size",
  title = "Average Transaction Value by Household Group Size",
  mean.ci = TRUE, 
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

## Significant Test of Association (Depedence) : *ggbarstats()* methods

### 1. Education Level

In the code chunk below, the Maths scores is binned into a 4-class variable by using [*cut()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/cut).

```{r}
customers_cleaned <- customers_cleaned %>% 
  mutate(Average_Transaction_bins = 
           cut(avg_tx_value, 
               breaks = c(0, 1339108, 4658605, Inf),
               labels = c("Low Spend", "Medium Spend", "High Spend"))
)
```

In this code chunk below [*ggbarstats()*](https://indrajeetpatil.github.io/ggstatsplot/reference/ggbarstats.html) is used to build a visual for Significant Test of Association

```{r}
ggbarstats(customers_cleaned, 
           x = Average_Transaction_bins, 
           y = education_level)
```

### 2. Marital Status

```{r}
ggbarstats(customers_cleaned, 
           x = Average_Transaction_bins, 
           y = marital_status)
```

### 3. Gender

```{r}
ggbarstats(customers_cleaned, 
           x = Average_Transaction_bins, 
           y = gender)
```

### 4. Age

```{r}
ggbarstats(customers_cleaned, 
           x = Average_Transaction_bins, 
           y = age_group)
```

### 5. Income Bracket

```{r}
ggbarstats(customers_cleaned, 
           x = Average_Transaction_bins, 
           y = income_bracket)
```

### 6. Household size

```{r}
ggbarstats(customers_cleaned, 
           x = Average_Transaction_bins, 
           y = hh_group)
```
